---
title: "Retrieving a Wikipedia Table: The Beatles' Studio Albums"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(wikitools)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
```

`wikitools` is built to facilitate the retrieval and manipulation of Wikipedia data. [as_wikitable()] converts an R data frame
*into* wikitext table markup for pasting into an article. [get_wikitable()] goes the other direction: it fetches the live, rendered HTML of a Wikipedia
article and pulls one of its `class="wikitable"` tables back out as a tidy
tibble.

This article walks through fetching one real table -- the studio-album
discography on [en.wikipedia.org/wiki/The_Beatles_albums_discography](https://en.wikipedia.org/wiki/The_Beatles_albums_discography)
-- and cleaning it up for analysis, which is most of the actual work: a
Wikipedia table's cells are HTML, not neatly typed data, so bullet lists,
footnote markers, and placeholder dashes all need to be dealt with before
anything downstream (a chart, a join, a summary) will work.

## Fetching the table

The page has several `wikitable`s -- one each for studio albums, live
albums, compilation albums, mash-ups, box sets, and EPs. `get_wikitable()`
picks one out by matching a (case-insensitive) substring against each
table's `<caption>`:

```{r}
albums <- get_wikitable(
  "The Beatles albums discography",
  match = "List of studio albums, with selected chart positions and certification"
)

albums
```

Three attributes travel along with the tibble: the caption actually
matched, the article URL that was fetched, and any single-cell "legend"
rows that were pulled out of the table body rather than kept as a data row:

```{r}
attr(albums, "caption")
attr(albums, "notes")
```

That note explains the `"\u2014"` placeholders scattered through the chart
columns below -- worth keeping in mind once those columns get converted to
numbers.

(One caveat surfaced while writing this: the page actually has *two*
captions containing "studio albums" -- a near-duplicate, much shorter table
further down covers a US-only release, "Introducing... The Beatles", that
isn't part of the main UK-catalog list above. `get_wikitable()` uses the
first match and prints a message when a `match` string is ambiguous like
this, so it's worth passing enough of the caption text to disambiguate, as
done above.)

## Header names and article links

By default, `get_wikitable()` also picks out any row-header column that
links to another article -- here, each album title links to its own page
-- and adds it as an adjacent `Title_link` column holding the absolute URL.
Ordinary cells (like the record label link buried inside `Album details`)
are left alone; only the row-header link is extracted, since it is the one
most likely to identify the row's own subject.

Two small helpers turn this into something more usable: `clean_column_headers()`
converts the raw header text into snake_case variable names (lower-casing
words but leaving all-caps acronyms like `UK`/`AUS` alone), and
`link_to_article_name()` turns a `_link` column's URL into a plain article
title:

```{r}
albums <- albums |>
  clean_column_headers() |>
  mutate(title_article = link_to_article_name(title_link)) |>
  relocate(title_article, .after = title)

albums |> select(title, title_article, title_link)
```

Worth a glance before moving on: the linked article title doesn't always
match the display title verbatim. *Revolver* links to the disambiguated
article "Revolver (Beatles album)", and "The Beatles (\"The White
Album\")" links to plain "The Beatles (album)" -- useful to know if
`title_article` is ever used as a join key against another data source.

## Choosing the right article interactively

`title_article` is a high-quality key because it reflects a Wikipedia
editor's deliberate choice of which article each album title links to. But
when a table *doesn't* link its rows (or when matching names against
Wikipedia from scratch), the usual fallback is search — and the top search
hit is not always the right article. Searching "A Hard Day's Night" puts
the film above the album; searching "With the Beatles" returns the band's
article; searching "Revolver" returns the firearm.

`choose_wikipedia_matches()` handles this interactively: for each name it
shows the top `n` search candidates (six by default) in a `utils::menu()`,
with each article's short description appended so similarly-titled articles
are easy to tell apart, plus a "None of these (skip)" option:

```{r eval = FALSE}
album_choices <- choose_wikipedia_matches(albums, name_col = "title")
```

```
Which article matches "Revolver"?

1: Revolver — Firearm with a cylinder holding cartridges
2: Revolver (Beatles album) — 1966 studio album by the Beatles
3: Revolver (disambiguation) — Topics referred to by the same term
...
7: None of these (skip)
```

The result is one row per query with `found`, `chosen`, the `rank` of the
selected candidate, and its `title`, `url`, and `description`.

### Making interactive choices reproducible

An interactive menu can't be part of a reproducible pipeline, but the
*choices* can. The pattern: run the picker once interactively, save the
result with `saveRDS()`, and have the script read the saved file on
subsequent (including non-interactive, e.g. knitted) runs:

```{r}
choices_file <- "album_choices.rds"

if (file.exists(choices_file)) {
  # Replay the previously recorded choices
  album_choices <- readRDS(choices_file)
} else if (interactive()) {
  album_choices <- choose_wikipedia_matches(albums, name_col = "title")
  saveRDS(album_choices, choices_file)
} else {
  stop("Run this script once interactively to record the choices.")
}

album_choices
```

The saved file is small and safe to commit alongside the script, so anyone
re-rendering this document gets exactly the choices that were made
interactively. (The `rank` column also documents how far down the list the
correct article was at the time the choices were made — search rankings do
drift over time, which is another reason to pin the result.)

### Comparing the two processes

The interactive choices can now be checked against what a fully automatic
top-hit search would have returned. Since the choices above were made using
the page's own links as the reference, `chosen_title` is the "right answer"
here — the question is how often [add_wikipedia_matches()] (which keeps
only the top search hit) lands on the same article:

Like the interactive choices, the automatic results are cached the first
time they're generated, so the comparison below stays fixed even as
Wikipedia's search rankings drift:

```{r}
auto_file <- "album_auto_matches.rds"

if (file.exists(auto_file)) {
  # Replay the previously recorded automatic matches
  album_auto_matches <- readRDS(auto_file)
} else if (interactive()) {
  album_auto_matches <- add_wikipedia_matches(albums, "title") |>
    select(title, automatic_title = wp_title, automatic_match = wp_match)
  saveRDS(album_auto_matches, auto_file)
} else {
  stop("Run this script once interactively to record the automatic matches.")
}

album_choices |>
  select(title = query, chosen_title = title, chosen_rank = rank) |>
  left_join(album_auto_matches, by = "title") |>
  mutate(same_article = chosen_title == automatic_title)
```

Nine of twelve agree. The three disagreements are the instructive ones:

- **"With the Beatles"** — the automatic top hit is the band's own article,
  not the album.
- **"A Hard Day's Night"** — the top hit is the *film*, not the album.
- **"Revolver"** — the top hit is the *firearm* article, and
  `automatic_match` is even `TRUE` (the title matches exactly!), so the
  exact-match flag would not have caught this. Only the short descriptions
  in the interactive picker make the firearm/album distinction visible.

Conversely, `automatic_match = FALSE` is not proof of a wrong result
either: for "The Beatles (\"The White Album\")", "Yellow Submarine", and
"Let It Be", the automatic search resolved to the correct disambiguated
article despite the display-title mismatch. The interactive step — or an
editor-curated link like `title_article` — is what turns these ambiguous
cases into confirmed matches.

## Cleaning the columns

Every column comes back as character data -- Wikipedia tables mix numbers,
dashes, and footnoted text too freely to safely guess types automatically.
The table has three groups of columns that each need different treatment:
`title` (and now `title_link`/`title_article`), plain text; the
chart-position columns (`UK`, `AUS`, `CAN`, `FRA`, `GER`, `NOR`, `US`),
which are almost numeric; and `album_details`, `certifications`, and
`sales`, which are really several lines of text packed into one cell.

### Chart positions

The `"\u2014"` placeholder (meaning "did not chart") needs to become `NA`
before these columns are usable as numbers:

```{r}
chart_cols <- c("UK", "AUS", "CAN", "FRA", "GER", "NOR", "US")

albums <- albums |>
  mutate(across(all_of(chart_cols), ~ na_if(.x, "\u2014"))) |>
  mutate(across(all_of(chart_cols), as.integer))

albums |> select(title, all_of(chart_cols))
```

### Release date and label

`album_details` packs a release date and a record label into one cell,
separated by a line break -- a pattern common to Wikipedia's `{{Plainlist}}`
and `<ul>` infobox-style cells:

```{r}
albums <- albums |>
  mutate(
    released = str_extract(album_details, "(?<=Released: )[^\\n]+") |> str_trim(),
    label    = str_extract(album_details, "(?<=Label: )[^\\n]+")    |> str_trim(),
    released = as.Date(released, format = "%d %B %Y")
  )

albums |> select(title, released, label)
```

### Certifications, reshaped long

`certifications` holds a variable number of `"Agency: Level"` lines per
album -- exactly the kind of one-cell-holds-many-values structure that
calls for `tidyr::separate_longer_delim()` followed by splitting each piece
on its colon:

```{r}
album_certifications <- albums |>
  select(title, certifications) |>
  filter(nzchar(certifications)) |>
  separate_longer_delim(certifications, delim = "\n") |>
  separate_wider_delim(certifications, delim = ": ", names = c("agency", "level")) |>
  mutate(agency = str_trim(agency), level = str_trim(level))

album_certifications
```

## A quick look: UK vs. US chart peaks over time

With `released` as a real date and the chart columns as real integers, a
release-order comparison of the two best-covered markets is a one-liner.
Remember that these are *chart positions*, so lower is better -- the axis
is flipped to read top-to-bottom as "higher on the chart" is "higher on the
plot":

```{r}
albums |>
  select(title, released, UK, US) |>
  pivot_longer(c(UK, US), names_to = "chart", values_to = "position") |>
  ggplot(aes(x = released, y = position, color = chart)) +
  geom_line(aes(group = chart)) +
  geom_point() +
  scale_y_reverse(breaks = c(1, 5, 10, 20, 40)) +
  labs(
    x = NULL, y = "Peak chart position", color = "Chart",
    title = "The Beatles: UK vs. US studio album chart peaks"
  )
```

Both charts show the same broad pattern -- every studio album from *Help!*
onward peaked at number one in the UK, while the US chart (which, unlike
the UK, treated several early releases as compilations rather than studio
albums, and so omits or reorders some titles) shows more variation early
on.

## Where this could go next

This is a first pass at the table, not a finished analysis. A few natural
extensions: joining `album_certifications` back to `albums` to compare
certification counts against chart performance; pulling in the other five
discography tables (live albums, compilations, etc.) with the same
`get_wikitable()` call and a different `match`; or cross-referencing
`title_article` against [add_wikidata_matches()] to attach each album's
Wikidata QID for further enrichment.
